---
Title: Data Mining Lab 2
Authors: Max Link, Logan Lu, Jadon Klipsch
Date: "2025-03-21"
Description: In this project, we will focus on clustering
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(repos=c(CRAN = "https://cloud.r-project.org"))

# Load required libraries!!
library(dplyr)      # For data manipulation
library(ggplot2)    # For visualizations
library(tidyr)      # For cleaning data
install.packages('ggrepel')
library(ggrepel) 
install.packages("xfun")


```


## 1.0 Problem Statement [0]

We will address the following key questions to aid Johnson and Johnson: 



1.	What are the counties in Texas with top ten infection rates? What are the measures that could be taken to reduce and tackle the infection rate? 

2.	How does median age impact infection and death rates? And what interventions could be taken to mitigate the risks for people with certain age groups?

3.	What key socioeconomic factors are most strongly associated with infection and death rates, and how can they inform targeted healthcare interventions?


## 2.0 Data Preparation [30]

```{r, results='hide', message=FALSE, warning=FALSE}
# Install all required packages (regardless of whether already installed)
install.packages(c("tidyverse", "factoextra", "cluster", "patchwork", "tibble", "ggrepel",
                   "mclust", "mcclust", "fpc", "seriation", "apcluster", "dbscan",
                   "entropy", "maps", "kernlab", "skimr", "readr", "dplyr", 
                   "tidyr", "knitr", "ggplot2"))
                   
```           

```{r, results='hide', message=FALSE, warning=FALSE}


# Load libraries
library(tidyverse)
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(tibble)
library(mclust)
library(mcclust)
library(ggplot2)
library(ggrepel)
library(cluster)
library(patchwork)
library(factoextra)
library(fpc)
library(seriation)
library(apcluster)
library(dbscan)
library(entropy)
library(maps)
library(skimr)


```

We can read in tx_data_with_gr.csv to get our same object from lab 1. 

```{r}

dataset <- read.csv("tx_data_with_gr.csv")

print(dataset)
```




1.	What are the counties in Texas with top ten infection rates? What are the measures that could be taken to reduce and tackle the infection rate? 

infection rate

county 



2.	How does median age impact infection and death rates? And what interventions could be taken to mitigate the risks for people with certain age groups?

3.	What key socioeconomic factors are most strongly associated with infection and death rates, and how can they inform targeted healthcare interventions?


We will cluster the following objects:


confirmed_cases (ratio) - clustering this variable will help us answer question 1 - 3 

median_age (ratio) - clustering this variable will answer question 2 

median_income (ratio) - clustering this variable will answer the 3rd question

total_pop (ratio) - clustering this variable will help us answer question 1 - 3 

rate_of_growth_first_30_days (ratio) - clustering this variable will help us answer question 1 

infection_rate (ratio) - clustering this variable will help answer question 1 - 3  

death_rate (ratio) - clustering this variable will help answer questions 1 - 3 

deaths(ratio) - clustering this variable will help answer questions 1 - 3 


Confirmed_Cases, total_pop, and median_income have larger numeric ranges than infection_rate and death_rate, so we need to standardize our variables 

We can use R's scale() function, which uses Z-score normalization

In Z-score normalization, values below the mean become negative and values above the mean become positive

```{r}

ds_scaled <- as.data.frame(scale(dataset[, c("confirmed_cases", "deaths", "median_age", "median_income", 
                          "total_pop", "rate_of_growth_first_30_days", 
                          "infection_rate", "death_rate")]))

# Add county_name back to the scaled dataset
ds_scaled$county_name <- dataset$county_name


```


We are using euclidian distance for our distance/similarity measurement 

## 3.0 Modeling [60] 


### 3.1 Question 1 - Infection Rate and County Clustering 


1. What are the counties in Texas with top ten infection rates? What are the measures that could be taken to reduce and tackle the infection rate? 

```{r}
# top ten counties by infection rate 

top10_infection <- ds_scaled[order(-ds_scaled$infection_rate), c("county_name", "infection_rate")][1:10, ]
print(top10_infection)



```

```{r}
# Subset the top 10 counties by infection rate
top10_infection <- ds_scaled[order(-ds_scaled$infection_rate), ][1:10, ]

# Select the variables for clustering
cluster_data <- top10_infection[, c("total_pop", "rate_of_growth_first_30_days", "confirmed_cases", "deaths", "infection_rate")]

# Verify the data
# print(cluster_data)

```

We can make an elbow plot to determine the number of clusters that are appropriate 

```{r}
# Load required libraries
library(ggplot2)

# Calculate WSS for different numbers of clusters (k from 1 to 8)
set.seed(123)  # For reproducibility
wss <- sapply(1:8, function(k) {
  kmeans(cluster_data, centers = k, nstart = 10)$tot.withinss
})

# Create a data frame for plotting
elbow_data <- data.frame(
  k = 1:8,
  WSS = wss
)

# Plot the elbow graph
ggplot(elbow_data, aes(x = k, y = WSS)) +
  geom_line() +
  geom_point() +
  labs(title = "Elbow Plot for Optimal Number of Clusters",
       x = "Number of Clusters (k)",
       y = "Within-Cluster Sum of Squares (WSS)") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:8)

```


```{r}

library(cluster)

# Calculate silhouette scores for different k
sil_scores <- sapply(2:8, function(k) {
  km <- kmeans(cluster_data, centers = k, nstart = 10)
  ss <- silhouette(km$cluster, dist(cluster_data))
  mean(ss[, 3])  # Average silhouette score
})

# Create a data frame for plotting
sil_data <- data.frame(
  k = 2:8,
  Silhouette = sil_scores
)

# Plot the silhouette scores
ggplot(sil_data, aes(x = k, y = Silhouette)) +
  geom_line() +
  geom_point() +
  labs(title = "Silhouette Plot for Optimal Number of Clusters",
       x = "Number of Clusters (k)",
       y = "Average Silhouette Score") +
  theme_minimal() +
  scale_x_continuous(breaks = 2:8)
```

The elbow plot suggests K = 3. The silhouette plot suggests K = 2. K = 3 can provide more tailored insights in our investigation because it enables better grouping for our county data (cities, rural areas, or both). 


We are looking at the top ten counties, which is a small sample of counties. K-means is appropriate for a small sample of counties because it is computationally efficient and is best used when the number of clusters is determined with an elbow or silhouette plot. 

```{r}

# Set seed for reproducibility
set.seed(123)

# Perform k-means clustering with 3 clusters
kmeans_result <- kmeans(cluster_data, centers = 3)

# Add cluster labels to the top10_infection dataset
top10_infection$cluster <- kmeans_result$cluster

# Display the results with county names and cluster assignments
result <- top10_infection[, c("county_name", "total_pop", "rate_of_growth_first_30_days", "confirmed_cases", "deaths", "infection_rate", "cluster")]
# print(result)

```


We can visualize our cluster results with ggplot 

```{r}
# Load required libraries
library(ggplot2)
library(ggrepel)  # For better label placement

# Step 1: Perform PCA on the clustering variables
# Extract the variables used for clustering (already scaled)
cluster_data <- top10_infection[, c("total_pop", "rate_of_growth_first_30_days", "confirmed_cases", "deaths", "infection_rate")]

# Run PCA
pca_result <- prcomp(cluster_data, scale. = FALSE)  # scale. = FALSE since data is already scaled

# Extract the first two principal components
pca_data <- as.data.frame(pca_result$x[, 1:2])  # PC1 and PC2
colnames(pca_data) <- c("PC1", "PC2")

# Add county names and cluster labels to the PCA data
pca_data$county_name <- top10_infection$county_name
pca_data$cluster <- as.factor(top10_infection$cluster)

# Step 2: Create the ggplot cluster plot
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster, label = county_name)) +
  geom_point(size = 4) +  # Plot points for each county
  geom_text_repel(size = 3, box.padding = 0.5, max.overlaps = Inf) +  # Add county names with ggrepel to avoid overlap
  scale_color_manual(values = c("red", "blue", "green")) +  # Color clusters
  labs(title = "Clustering of Top 10 Counties by Infection Rate",
       x = "Principal Component 1",
       y = "Principal Component 2",
       color = "Cluster") +
  theme_minimal() +
  theme(legend.position = "bottom")

```


